Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-10-14 article.tex
Addressed point 19
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1404 ccdd05d5-56c1-461a-bf98-85f6876523b4
Addressed point 2
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1405 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW\
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\NS) is involved, it might become tidally disrupted near the
merger and fuel an electromagnetic (\EM) counterpart \citep{shibata:2007}.
Effort from both the \GW\ and the broader astronomical communities might make
it possible to use \GW\ observations as early warning triggers for \EM\
followup. In the first generation of ground-based laser interferometers, the
\GW\ community initiated a project to send alerts when potential \GW\
transients were observed in order to trigger followup observations by \EM\
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011}.
This was an important achievement, but too late to catch any prompt \EM\
emission.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one might have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of
this problem, a computationally inexpensive latency free, real-time filtering
algorithm for detecting inspiral signals in \GW\ data.  We also consider the
prospects for advanced \GW\ detectors and discuss other areas of work that would
be required for rapid analysis.

@@

\CBC\ events may be the progenitors of some short hard \GRB{}s and are expected
to be accompanied by a broad spectrum of \EM\ signals. Rapid alerts to the
wider astronomical community will improve the chances of detecting an \EM\
counterpart in bands from gamma-rays down to radio. In the Advanced LIGO
era, it appears possible to usefully localize a few rare events prior to the
\GRB{}, allowing multi-wavelength observations of prompt emission. More
frequently, low-latency alerts will be released after merger but will still
yield extended X-ray tails and early afterglows.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW\
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\NS) is involved, it might become tidally disrupted near the
merger and fuel an electromagnetic (\EM) counterpart \citep{shibata:2007}.
Effort from both the \GW\ and the broader astronomical communities might make
it possible to use \GW\ observations as early warning triggers for \EM\
followup. In the first generation of ground-based laser interferometers, the
\GW\ community initiated a project to send alerts when potential \GW\
transients were observed in order to trigger followup observations by \EM\
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011}.
This was an important achievement, but too late to catch any prompt optical flash
or the onset of an on-axis optical afterglow.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one might have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of
this problem, a computationally inexpensive latency free, real-time filtering
algorithm for detecting inspiral signals in \GW\ data.  We also consider the
prospects for advanced \GW\ detectors and discuss other areas of work that would
be required for rapid analysis.

@@

\CBC\ events may be the progenitors of some short hard \GRB{}s and are expected
to be accompanied by a broad spectrum of \EM\ signals. Rapid alerts to the
wider astronomical community will improve the chances of detecting an \EM\
counterpart in bands from gamma-rays down to radio. In the Advanced LIGO
era, it appears possible to usefully localize a few rare events prior to the
\GRB{}, allowing multi-wavelength observations of prompt emission. More
frequently, low-latency alerts will be released after merger but may still
yield extended X-ray tails and early on-axis afterglows.
